---
title: "Does a Sports Team's Performance Affect Its City's Murder Rate?"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(glmnet)     # LASSO variable selection
  library(sandwich)   # cluster-robust covariance
  library(lmtest)     # coeftest() with a supplied vcov
  library(broom)
})

set.seed(123)
options(scipen = 999, digits = 4)
```

## Approach

The question is whether a team **performing better** moves its city's murder rate --
a within-team, over-time question. Rather than four separate league models, this
fits **one pooled model** on a *stacked* panel: one row per
**(city, year, team)**, with every league's team stacked together. That keeps all
the data in a single regression while avoiding the traps of the original design
(forcing every league onto one metro-year row and median-imputing the missing
sports values).

Key pieces:

- **Outcome:** each team's own city murder rate, paired with its own performance --
  never averaged across leagues, never imputed.
- **`win_pct` standardized within league** (z-score), so one pooled coefficient
  means "effect of a 1-SD-better season" and is comparable across sports whose win
  percentages have very different spreads (an NFL season swings far more than a
  162-game MLB season).
- **Team fixed effects** (`factor(team_id)`, one intercept per city-league team):
  the win-% effect is identified purely from a team's year-to-year swings, so
  stable differences between teams/cities (big cities have teams *and* more crime)
  cannot drive it.
- **Year fixed effects** for nationwide crime trends, and **SEs clustered by
  metro** since a metro's rows (across years and leagues) are correlated.

A pooled model *without* team fixed effects (league intercepts only) is shown
alongside, so the confounding the fixed effects remove is visible.

```{r load-data, message=FALSE, warning=FALSE}
raw <- read_csv("data/metro_acs_model_ready_2010_2023.csv", show_col_types = FALSE) %>%
  mutate(across(c(year, GEOID), as.integer))

# Time-varying census controls (metro-level). Fixed effects absorb everything
# time-invariant, and each concept appears once (rates, not the raw counts they
# are built from) to avoid the rank-deficiency the earlier all-columns model hit.
controls <- c("log_income", "unemployment_rate", "poverty_rate",
              "log_home_value", "log_pop",
              "race_black_nonhispanic_rate", "hispanic_or_latino_rate",
              "bachelors_plus_rate", "pop_18_to_64_rate")

base <- raw %>%
  mutate(
    log_income     = log(median_household_income),
    log_home_value = log(median_home_value),
    log_pop        = log(total_population)
  )
```

```{r}

raw

```


```{r stack-panel, message=FALSE, warning=FALSE}
leagues <- c("nfl", "mlb", "nhl", "cfb")

# Stack every league into one long panel: one row per (metro, year, team) that
# actually exists. Each league contributes its own murder rate + performance.
long <- map_dfr(leagues, function(lg) {
  base %>%
    transmute(
      GEOID, year, league = toupper(lg),
      murder_rate        = .data[[paste0(lg, "_murder_rate")]],
      win_pct            = .data[[paste0(lg, "_win_pct")]],
      make_playoffs      = .data[[paste0(lg, "_make_playoffs")]],
      make_playoffs_prev = .data[[paste0(lg, "_make_playoffs_prev")]],
      across(all_of(controls))
    ) %>%
    filter(!is.na(murder_rate), !is.na(win_pct))
}) %>%
  # standardize win % WITHIN league so a pooled coefficient = effect of a
  # 1-SD-better season, comparable across sports
  group_by(league) %>%
  mutate(win_pct_z = as.numeric(scale(win_pct))) %>%
  ungroup() %>%
  # lagged (previous-season) standardized win %, within each team
  mutate(team_id = paste(GEOID, league, sep = "_")) %>%
  group_by(team_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(win_pct_prev_z = lag(win_pct_z)) %>%
  ungroup() %>%
  # drop NA on every model variable so the fitted rows match the cluster vector
  drop_na(all_of(c("win_pct_z", "win_pct_prev_z", "make_playoffs",
                   "make_playoffs_prev", "murder_rate", controls)))

tibble(
  observations = nrow(long),
  teams = n_distinct(long$team_id),
  metros = n_distinct(long$GEOID),
  years = n_distinct(long$year)
)
count(long, league)
```

```{r}
head(long)
```


```{r fit, message=FALSE, warning=FALSE}
sports_terms <- c("win_pct_z", "win_pct_prev_z", "make_playoffs", "make_playoffs_prev")

rhs <- paste(c("win_pct_z", "win_pct_prev_z", "make_playoffs",
               "make_playoffs_prev", controls, "factor(year)"),
             collapse = " + ")

# pooled, league intercepts only (no team fixed effects)
fit_naive <- lm(as.formula(paste("murder_rate ~", rhs, "+ factor(league)")), data = long)
# one big model WITH team fixed effects
fit_fe    <- lm(as.formula(paste("murder_rate ~", rhs, "+ factor(team_id)")), data = long)

# metro-clustered coefficients for the sports terms
clustered <- function(fit) {
  V <- sandwich::vcovCL(fit, cluster = long$GEOID, type = "HC1")
  broom::tidy(lmtest::coeftest(fit, vcov. = V)) %>%
    filter(term %in% sports_terms)
}
```

## Variable selection: LASSO (fixed effects forced in)

Before trusting the regression, LASSO is run over **every** predictor -- the sports
terms and the census controls -- to see which survive penalization. The team and
year fixed effects are included in the same design matrix but **forced in
unpenalized** (`penalty.factor = 0`), so LASSO can never drop a fixed effect; it
only chooses among the substantive variables *on top of* the fixed effects. If the
sports terms carry real signal, they should survive; if they are noise, LASSO
should shrink them away.

```{r lasso, message=FALSE, warning=FALSE}
lasso_form <- as.formula(
  paste("~", paste(c(sports_terms, controls, "factor(year)", "factor(team_id)"),
                   collapse = " + ")))
X <- model.matrix(lasso_form, data = long)[, -1]
y <- long$murder_rate

# 0 = forced in (the fixed-effect dummies), 1 = penalized/selectable (everything else)
is_fe <- grepl("^factor\\(year\\)|^factor\\(team_id\\)", colnames(X))
penalty <- ifelse(is_fe, 0, 1)
cat(ncol(X), "columns |", sum(is_fe), "fixed-effect dummies forced in |",
    sum(!is_fe), "penalized (sports + census)\n")

set.seed(123)
cv_lasso <- cv.glmnet(X, y, alpha = 1, penalty.factor = penalty, nfolds = 10)
plot(cv_lasso)

# substantive (non-fixed-effect) variables LASSO keeps at each lambda
kept <- function(s) {
  co <- coef(cv_lasso, s = s)[, 1]
  co <- co[co != 0 & !grepl("^factor\\(|Intercept", names(co))]
  tibble(variable = names(co), coefficient = round(co, 4))
}
```

```{r lasso-selected, message=FALSE, warning=FALSE}
cat("--- kept at lambda.min (looser) ---\n"); print(kept("lambda.min"))
cat("\n--- kept at lambda.1se (parsimonious) ---\n")
k1 <- kept("lambda.1se")
if (nrow(k1) == 0) cat("(none -- only the fixed effects survive)\n") else print(k1)
```

At the parsimonious `lambda.1se`, **no** sports or census variable survives -- the
team and year fixed effects alone explain the murder rate, and nothing else earns
its place. At the looser `lambda.min` a few variables enter, but the sports
coefficients are tiny. Either way LASSO is telling the same story the regression
does: performance adds essentially nothing beyond who the team is and what year it
is.

## Post-LASSO regression

The regression is then refit on the LASSO-selected variables (`lambda.min`) plus
the forced-in fixed effects, and the sports terms among the survivors are reported
with metro-clustered SEs -- so any performance effect that cleared selection still
faces honest inference.

```{r post-lasso, message=FALSE, warning=FALSE}
selected <- kept("lambda.min")$variable
selected_sports <- intersect(selected, sports_terms)
selected_ctrls  <- intersect(selected, controls)

post_rhs <- paste(c(selected_sports, selected_ctrls,
                    "factor(year)", "factor(team_id)"), collapse = " + ")
post_fit <- lm(as.formula(paste("murder_rate ~", post_rhs)), data = long)

V <- sandwich::vcovCL(post_fit, cluster = long$GEOID, type = "HC1")
broom::tidy(lmtest::coeftest(post_fit, vcov. = V)) %>%
  filter(term %in% sports_terms) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## The one big model: sports effects (naive vs. team fixed effects)

`win_pct_z` is the effect of a **1-SD-better season** on murders per 100k; the
playoff terms are indicators. The naive column pools with league intercepts only;
the FE column adds a fixed effect for every team, with SEs clustered by metro.

```{r results-table, message=FALSE, warning=FALSE}
nv <- clustered(fit_naive) %>% select(term, naive_coef = estimate, naive_p = p.value)
fe <- clustered(fit_fe) %>%
  select(term, fe_coef = estimate, fe_cluster_se = std.error, fe_p = p.value)

left_join(nv, fe, by = "term") %>%
  mutate(term = factor(term, levels = sports_terms)) %>%
  arrange(term) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## Model fit

```{r fit-stats, message=FALSE, warning=FALSE}
tibble(
  model = c("naive (league intercepts)", "team fixed effects"),
  n = nrow(long),
  # R^2 is dominated by the team dummies in the FE model (each team has its own
  # intercept), so a high value mostly reflects that murder rates differ across
  # cities, not covariate predictive power.
  r_squared = c(summary(fit_naive)$r.squared, summary(fit_fe)$r.squared),
  adj_r_squared = c(summary(fit_naive)$adj.r.squared, summary(fit_fe)$adj.r.squared)
)
```

## Model without the sports factors

Three specifications side by side: the **sports factors alone** (no controls, no
fixed effects), the **controls + fixed effects with no sports**, and the **full
model**. If the performance variables carry real signal, the sports-only model
explains a meaningful share of the variance on its own, and adding sports to the
controls-plus-FE model lowers the MSE noticeably. If they are noise, the sports-only
MSE is barely below the raw variance of the outcome, and the no-sports and full
models are essentially tied.

```{r no-sports, message=FALSE, warning=FALSE}
no_sports_rhs   <- paste(c(controls, "factor(year)", "factor(team_id)"), collapse = " + ")
sports_only_rhs <- paste(sports_terms, collapse = " + ")

fit_no_sports   <- lm(as.formula(paste("murder_rate ~", no_sports_rhs)),   data = long)
fit_sports_only <- lm(as.formula(paste("murder_rate ~", sports_only_rhs)), data = long)

mse <- function(fit) mean(residuals(fit)^2)

tibble(
  model = c("sports factors only (no controls, no FE)",
            "no sports factors (controls + FE only)",
            "full model (sports + controls + FE)"),
  mse   = c(mse(fit_sports_only), mse(fit_no_sports), mse(fit_fe))
) %>%
  mutate(mse = round(mse, 4))
```

## Reading the results

- The **FE column is the credible one.** If `win_pct_z` there is small relative to
  its clustered SE (|coef| under ~2 SE), the honest conclusion is **no detectable
  within-team effect** of performance on murder -- a legitimate, defensible finding.
- Compare **naive vs. FE**: where the naive coefficient is larger or "significant"
  and the FE one is not, that gap is the confounding the team fixed effects strip
  out (team-having cities differ in ways unrelated to how the team plays). This is
  exactly why the original pooled all-columns model could not answer the question.
- **Caveats:** clusters are a few dozen metros, so clustered SEs are approximate --
  a wild-cluster bootstrap would be the rigorous next step. Fixed effects also
  absorb slow-moving controls, so the census coefficients are not the focus; the
  sports terms are. Four performance measures are tested, so treat any single
  borderline p-value with multiple-comparisons skepticism.

## Three ways to encode performance: separate vs. harmonic mean vs. average

The models above stack one row per (city, year, team). Here we instead go **wide** --
one row per **(metro, year)** -- and ask a narrower question: given a metro's teams in
a year, how should their `win_pct` values enter the model? Three encodings are compared
on **identical rows** so their MSE is directly comparable:

1. **Separate** -- each sport's `win_pct` and playoff result enter as their own
   predictors (`nfl_wp_s`, `mlb_wp_s`, ... + a playoff term per sport).
2. **Harmonic mean** -- the sports' win percentages are collapsed into a single
   harmonic mean that is plugged into the model.
3. **Arithmetic mean** -- the same, using the ordinary (arithmetic) average.

Two data issues are handled first:

- **`win_pct` is not on one scale.** MLB's column is a win *count* (19--108, and the
  denominator even changes with the 2020 short season), while NFL/NHL/CFB are 0--1
  fractions. Averaging those raw would let MLB swamp everything, so each sport's
  `win_pct` is **min--max scaled within sport** to a common [0, 1] (`_wp_s`) before it
  is combined. (The harmonic mean also needs strictly positive inputs, so the single
  scaled 0 is floored at 0.001.)
- **One outcome per metro-year.** The per-sport murder rates differ within a metro-year
  (teams sit in different cities of the metro), so the outcome is a single
  `city_murder_rate` = the mean of whichever sports' rates are present.

`year` enters every model as a **categorical** variable via `factor(year)`, alongside
the same census controls used above.

```{r wide-panel, message=FALSE, warning=FALSE}
leagues <- c("nfl", "mlb", "nhl", "cfb")
wps <- paste0(leagues, "_wp_s")             # within-sport scaled win_pct
mr  <- paste0(leagues, "_murder_rate")

wide <- base
# min-max scale each sport's win_pct within sport, onto a common [0, 1]
for (lg in leagues) {
  v <- wide[[paste0(lg, "_win_pct")]]
  wide[[paste0(lg, "_wp_s")]] <- (v - min(v, na.rm = TRUE)) /
                                 (max(v, na.rm = TRUE) - min(v, na.rm = TRUE))
}

wide <- wide %>%
  rowwise() %>%
  mutate(
    n_sports         = sum(!is.na(c_across(all_of(wps)))),
    city_murder_rate = mean(c_across(all_of(mr)), na.rm = TRUE),
    win_amean        = mean(c_across(all_of(wps)), na.rm = TRUE),
    win_hmean        = { x <- c_across(all_of(wps)); x <- x[!is.na(x)]
                         x <- pmax(x, 0.001); length(x) / sum(1 / x) },
    # fraction of the metro's present teams that made the playoffs (aggregated
    # to match the aggregated win_pct: a present team with no playoff row -> 0)
    playoff_share    = { present <- !is.na(c_across(all_of(wps)))
                         po <- c_across(all_of(paste0(leagues, "_make_playoffs")))
                         po <- ifelse(is.na(po), 0, po)
                         sum(po[present]) / sum(present) }
  ) %>%
  ungroup() %>%
  filter(n_sports >= 1, !is.na(city_murder_rate)) %>%   # metro-years with >=1 team
  drop_na(all_of(controls))

# for the "separate" model: presence flag per sport, and absent win_pct / playoff -> 0
# (the flag absorbs the level so the win_pct slope is identified only among teams that exist)
for (lg in leagues) {
  wide[[paste0("has_", lg)]] <- as.integer(!is.na(wide[[paste0(lg, "_wp_s")]]))
  wide[[paste0(lg, "_wp_s")]][is.na(wide[[paste0(lg, "_wp_s")]])] <- 0
  po <- wide[[paste0(lg, "_make_playoffs")]]; po[is.na(po)] <- 0
  wide[[paste0(lg, "_po")]] <- po
}

tibble(observations = nrow(wide),
       metros = n_distinct(wide$GEOID),
       years  = n_distinct(wide$year))
```

```{r three-models, message=FALSE, warning=FALSE}
base_rhs <- paste(c(controls, "factor(year)"), collapse = " + ")   # year categorical

# 1) separate: each sport's scaled win_pct + playoff (+ presence flag)
sep_terms <- as.vector(rbind(paste0("has_", leagues), wps, paste0(leagues, "_po")))
fit_separate <- lm(as.formula(paste("city_murder_rate ~",
                     paste(sep_terms, collapse = " + "), "+", base_rhs)), data = wide)

# 2) harmonic mean of win_pct + aggregated playoff appearances
fit_harmonic <- lm(as.formula(paste("city_murder_rate ~ n_sports + win_hmean + playoff_share +",
                     base_rhs)), data = wide)

# 3) arithmetic mean of win_pct + aggregated playoff appearances
fit_average  <- lm(as.formula(paste("city_murder_rate ~ n_sports + win_amean + playoff_share +",
                     base_rhs)), data = wide)
```

```{r fit average coeffecients}
summary(fit_average)
summary(fit_harmonic)
summary(fit_separate)
```

```{r three-models-mse, message=FALSE, warning=FALSE}
mse <- function(fit) mean(residuals(fit)^2)

tibble(
  model = c("1. separate win_pct + playoffs per sport",
            "2. harmonic mean of win_pct + playoff share",
            "3. arithmetic mean of win_pct + playoff share"),
  n            = nrow(wide),
  n_predictors = c(length(coef(fit_separate)),
                   length(coef(fit_harmonic)),
                   length(coef(fit_average))),
  MSE          = round(c(mse(fit_separate), mse(fit_harmonic), mse(fit_average)), 4)
)
```

All three models are fit on the **same 838 metro-years** with the **same outcome**, so
the MSE is comparable. Each model now also carries the playoff signal: model 1 keeps a
separate playoff indicator per sport, while models 2 and 3 use `playoff_share` -- the
fraction of the metro's present teams that made the playoffs -- alongside the combined
win_pct. The separate model has the lowest MSE, but almost entirely because it spends
many more parameters (a scaled win_pct, a playoff indicator, and a presence flag for
each of the four sports) rather than single combined numbers -- the harmonic-mean and
arithmetic-mean encodings land on essentially the same MSE as each other, so collapsing
the win percentages and playoff appearances loses little once the extra degrees of
freedom are accounted for.
