---
title: "Does a Sports Team's Performance Affect Its City's Murder Rate?"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(glmnet)     # LASSO variable selection
  library(sandwich)   # cluster-robust covariance
  library(lmtest)     # coeftest() with a supplied vcov
  library(broom)
})

set.seed(123)
options(scipen = 999, digits = 4)
```

## Approach

The question is whether a team **performing better** moves its city's murder rate --
a within-team, over-time question. Rather than four separate league models, this
fits **one pooled model** on a *stacked* panel: one row per
**(city, year, team)**, with every league's team stacked together. That keeps all
the data in a single regression while avoiding the traps of the original design
(forcing every league onto one metro-year row and median-imputing the missing
sports values).

Key pieces:

- **Outcome:** each team's own city murder rate, paired with its own performance --
  never averaged across leagues, never imputed.
- **`win_pct` standardized within league** (z-score), so one pooled coefficient
  means "effect of a 1-SD-better season" and is comparable across sports whose win
  percentages have very different spreads (an NFL season swings far more than a
  162-game MLB season).
- **Team fixed effects** (`factor(team_id)`, one intercept per city-league team):
  the win-% effect is identified purely from a team's year-to-year swings, so
  stable differences between teams/cities (big cities have teams *and* more crime)
  cannot drive it.
- **Year fixed effects** for nationwide crime trends, and **SEs clustered by
  metro** since a metro's rows (across years and leagues) are correlated.

A pooled model *without* team fixed effects (league intercepts only) is shown
alongside, so the confounding the fixed effects remove is visible.

```{r load-data, message=FALSE, warning=FALSE}
raw <- read_csv("data/metro_acs_model_ready_2010_2023.csv", show_col_types = FALSE) %>%
  mutate(across(c(year, GEOID), as.integer))

# Time-varying census controls (metro-level). Fixed effects absorb everything
# time-invariant, and each concept appears once (rates, not the raw counts they
# are built from) to avoid the rank-deficiency the earlier all-columns model hit.
controls <- c("log_income", "unemployment_rate", "poverty_rate",
              "log_home_value", "log_pop",
              "race_black_nonhispanic_rate", "hispanic_or_latino_rate",
              "bachelors_plus_rate", "pop_18_to_64_rate")

base <- raw %>%
  mutate(
    log_income     = log(median_household_income),
    log_home_value = log(median_home_value),
    log_pop        = log(total_population)
  )
```

```{r}

raw

```


```{r stack-panel, message=FALSE, warning=FALSE}
leagues <- c("nfl", "mlb", "nhl", "cfb")

# Stack every league into one long panel: one row per (metro, year, team) that
# actually exists. Each league contributes its own murder rate + performance.
long <- map_dfr(leagues, function(lg) {
  base %>%
    transmute(
      GEOID, year, league = toupper(lg),
      murder_rate        = .data[[paste0(lg, "_murder_rate")]],
      win_pct            = .data[[paste0(lg, "_win_pct")]],
      make_playoffs      = .data[[paste0(lg, "_make_playoffs")]],
      make_playoffs_prev = .data[[paste0(lg, "_make_playoffs_prev")]],
      across(all_of(controls))
    ) %>%
    filter(!is.na(murder_rate), !is.na(win_pct))
}) %>%
  # standardize win % WITHIN league so a pooled coefficient = effect of a
  # 1-SD-better season, comparable across sports
  group_by(league) %>%
  mutate(win_pct_z = as.numeric(scale(win_pct))) %>%
  ungroup() %>%
  # lagged (previous-season) standardized win %, within each team
  mutate(team_id = paste(GEOID, league, sep = "_")) %>%
  group_by(team_id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(win_pct_prev_z = lag(win_pct_z)) %>%
  ungroup() %>%
  # drop NA on every model variable so the fitted rows match the cluster vector
  drop_na(all_of(c("win_pct_z", "win_pct_prev_z", "make_playoffs",
                   "make_playoffs_prev", "murder_rate", controls)))

tibble(
  observations = nrow(long),
  teams = n_distinct(long$team_id),
  metros = n_distinct(long$GEOID),
  years = n_distinct(long$year)
)
count(long, league)
```

```{r fit, message=FALSE, warning=FALSE}
sports_terms <- c("win_pct_z", "win_pct_prev_z", "make_playoffs", "make_playoffs_prev")

rhs <- paste(c("win_pct_z", "win_pct_prev_z", "make_playoffs",
               "make_playoffs_prev", controls, "factor(year)"),
             collapse = " + ")

# pooled, league intercepts only (no team fixed effects)
fit_naive <- lm(as.formula(paste("murder_rate ~", rhs, "+ factor(league)")), data = long)
# one big model WITH team fixed effects
fit_fe    <- lm(as.formula(paste("murder_rate ~", rhs, "+ factor(team_id)")), data = long)

# metro-clustered coefficients for the sports terms
clustered <- function(fit) {
  V <- sandwich::vcovCL(fit, cluster = long$GEOID, type = "HC1")
  broom::tidy(lmtest::coeftest(fit, vcov. = V)) %>%
    filter(term %in% sports_terms)
}
```

## Variable selection: LASSO (fixed effects forced in)

Before trusting the regression, LASSO is run over **every** predictor -- the sports
terms and the census controls -- to see which survive penalization. The team and
year fixed effects are included in the same design matrix but **forced in
unpenalized** (`penalty.factor = 0`), so LASSO can never drop a fixed effect; it
only chooses among the substantive variables *on top of* the fixed effects. If the
sports terms carry real signal, they should survive; if they are noise, LASSO
should shrink them away.

```{r lasso, message=FALSE, warning=FALSE}
lasso_form <- as.formula(
  paste("~", paste(c(sports_terms, controls, "factor(year)", "factor(team_id)"),
                   collapse = " + ")))
X <- model.matrix(lasso_form, data = long)[, -1]
y <- long$murder_rate

# 0 = forced in (the fixed-effect dummies), 1 = penalized/selectable (everything else)
is_fe <- grepl("^factor\\(year\\)|^factor\\(team_id\\)", colnames(X))
penalty <- ifelse(is_fe, 0, 1)
cat(ncol(X), "columns |", sum(is_fe), "fixed-effect dummies forced in |",
    sum(!is_fe), "penalized (sports + census)\n")

set.seed(123)
cv_lasso <- cv.glmnet(X, y, alpha = 1, penalty.factor = penalty, nfolds = 10)
plot(cv_lasso)

# substantive (non-fixed-effect) variables LASSO keeps at each lambda
kept <- function(s) {
  co <- coef(cv_lasso, s = s)[, 1]
  co <- co[co != 0 & !grepl("^factor\\(|Intercept", names(co))]
  tibble(variable = names(co), coefficient = round(co, 4))
}
```

```{r lasso-selected, message=FALSE, warning=FALSE}
cat("--- kept at lambda.min (looser) ---\n"); print(kept("lambda.min"))
cat("\n--- kept at lambda.1se (parsimonious) ---\n")
k1 <- kept("lambda.1se")
if (nrow(k1) == 0) cat("(none -- only the fixed effects survive)\n") else print(k1)
```

At the parsimonious `lambda.1se`, **no** sports or census variable survives -- the
team and year fixed effects alone explain the murder rate, and nothing else earns
its place. At the looser `lambda.min` a few variables enter, but the sports
coefficients are tiny. Either way LASSO is telling the same story the regression
does: performance adds essentially nothing beyond who the team is and what year it
is.

## Post-LASSO regression

The regression is then refit on the LASSO-selected variables (`lambda.min`) plus
the forced-in fixed effects, and the sports terms among the survivors are reported
with metro-clustered SEs -- so any performance effect that cleared selection still
faces honest inference.

```{r post-lasso, message=FALSE, warning=FALSE}
selected <- kept("lambda.min")$variable
selected_sports <- intersect(selected, sports_terms)
selected_ctrls  <- intersect(selected, controls)

post_rhs <- paste(c(selected_sports, selected_ctrls,
                    "factor(year)", "factor(team_id)"), collapse = " + ")
post_fit <- lm(as.formula(paste("murder_rate ~", post_rhs)), data = long)

V <- sandwich::vcovCL(post_fit, cluster = long$GEOID, type = "HC1")
broom::tidy(lmtest::coeftest(post_fit, vcov. = V)) %>%
  filter(term %in% sports_terms) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## The one big model: sports effects (naive vs. team fixed effects)

`win_pct_z` is the effect of a **1-SD-better season** on murders per 100k; the
playoff terms are indicators. The naive column pools with league intercepts only;
the FE column adds a fixed effect for every team, with SEs clustered by metro.

```{r results-table, message=FALSE, warning=FALSE}
nv <- clustered(fit_naive) %>% select(term, naive_coef = estimate, naive_p = p.value)
fe <- clustered(fit_fe) %>%
  select(term, fe_coef = estimate, fe_cluster_se = std.error, fe_p = p.value)

left_join(nv, fe, by = "term") %>%
  mutate(term = factor(term, levels = sports_terms)) %>%
  arrange(term) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## Model fit

```{r fit-stats, message=FALSE, warning=FALSE}
tibble(
  model = c("naive (league intercepts)", "team fixed effects"),
  n = nrow(long),
  # R^2 is dominated by the team dummies in the FE model (each team has its own
  # intercept), so a high value mostly reflects that murder rates differ across
  # cities, not covariate predictive power.
  r_squared = c(summary(fit_naive)$r.squared, summary(fit_fe)$r.squared),
  adj_r_squared = c(summary(fit_naive)$adj.r.squared, summary(fit_fe)$adj.r.squared)
)
```

## Model without the sports factors

Three specifications side by side: the **sports factors alone** (no controls, no
fixed effects), the **controls + fixed effects with no sports**, and the **full
model**. If the performance variables carry real signal, the sports-only model
explains a meaningful share of the variance on its own, and adding sports to the
controls-plus-FE model lowers the MSE noticeably. If they are noise, the sports-only
MSE is barely below the raw variance of the outcome, and the no-sports and full
models are essentially tied.

```{r no-sports, message=FALSE, warning=FALSE}
no_sports_rhs   <- paste(c(controls, "factor(year)", "factor(team_id)"), collapse = " + ")
sports_only_rhs <- paste(sports_terms, collapse = " + ")

fit_no_sports   <- lm(as.formula(paste("murder_rate ~", no_sports_rhs)),   data = long)
fit_sports_only <- lm(as.formula(paste("murder_rate ~", sports_only_rhs)), data = long)

mse <- function(fit) mean(residuals(fit)^2)

tibble(
  model = c("sports factors only (no controls, no FE)",
            "no sports factors (controls + FE only)",
            "full model (sports + controls + FE)"),
  mse   = c(mse(fit_sports_only), mse(fit_no_sports), mse(fit_fe))
) %>%
  mutate(mse = round(mse, 4))
```

## Reading the results

- The **FE column is the credible one.** If `win_pct_z` there is small relative to
  its clustered SE (|coef| under ~2 SE), the honest conclusion is **no detectable
  within-team effect** of performance on murder -- a legitimate, defensible finding.
- Compare **naive vs. FE**: where the naive coefficient is larger or "significant"
  and the FE one is not, that gap is the confounding the team fixed effects strip
  out (team-having cities differ in ways unrelated to how the team plays). This is
  exactly why the original pooled all-columns model could not answer the question.
- **Caveats:** clusters are a few dozen metros, so clustered SEs are approximate --
  a wild-cluster bootstrap would be the rigorous next step. Fixed effects also
  absorb slow-moving controls, so the census coefficients are not the focus; the
  sports terms are. Four performance measures are tested, so treat any single
  borderline p-value with multiple-comparisons skepticism.









Team + year fixed effects. That means every census coefficient is identified only from within-metro, year-to-year deviations

"what is most correlated with crime"


"Does changing X move murder within a metro?"


```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(glmnet)     # LASSO variable selection
  library(sandwich)   # cluster-robust covariance
  library(lmtest)     # coeftest() with a supplied vcov
  library(broom)
})

set.seed(123)
options(scipen = 999, digits = 4)
```


```{r correlates-clean, message=FALSE, warning=FALSE}
raw <- read_csv("data/metro_acs_model_ready_2010_2023.csv", show_col_types = FALSE) %>%
  mutate(across(c(year, GEOID), as.integer))
```


```{r}
base <- raw %>%
  mutate(
    log_income     = log(median_household_income),
    log_home_value = log(median_home_value),
    log_pop        = log(total_population)
  )
```

```{r}
race_cols <- c("race_white_nonhispanic_rate","race_black_nonhispanic_rate",
               "race_aian_nonhispanic_rate","race_asian_nonhispanic_rate",
               "race_nhpi_nonhispanic_rate","race_other_nonhispanic_rate",
               "race_two_plus_nonhispanic_rate","hispanic_or_latino_rate")

    
disadv_vars <- c("poverty_rate","unemployment_rate","log_income",
                 "bachelors_plus_rate","race_black_nonhispanic_rate")
```

```{r}
acs <- base %>%
  mutate(
    murder_rate   = rowMeans(across(ends_with("_murder_rate")), na.rm = TRUE),
    ethnic_hetero = 1 - rowSums(across(all_of(race_cols))^2),  # Blau index
    renter_share  = 1 - homeownership_rate,
    youth_share   = pop_under_18_rate
  ) %>%
  filter(!is.nan(murder_rate))
```
murder_rate = rowMeans(across(ends_with("_murder_rate")), na.rm = TRUE)
Takes all columns whose names end in _murder_rate and averages them row by row. So if a metro-year has multiple league-specific murder-rate columns, this creates one overall murder-rate measure per row.

ethnic_hetero = 1 - rowSums(across(all_of(race_cols))^2)
Computes an ethnic heterogeneity index, also called a Blau index. If the race shares are more evenly spread out, this value is higher. If one group dominates, it is lower.

renter_share = 1 - homeownership_rate
Converts homeownership into renter share.

youth_share = pop_under_18_rate
Just renames/copies the under-18 population share into a shorter variable name.

filter(!is.nan(murder_rate))
Drops rows where murder_rate is NaN, which can happen if all the _murder_rate columns were missing for that row.

```{r}
# concentrated disadvantage = PC1 of the standardized, collinear SES block
# why? log_income (values ~11) would dominate unemployment_rate (values ~0.05) just because its numbers are bigger
Xpc   <- scale(as.matrix(acs[disadv_vars]))
cc    <- complete.cases(Xpc)
pc    <- prcomp(Xpc[cc, , drop = FALSE])
score <- rep(NA_real_, nrow(acs)); score[cc] <- pc$x[, 1]
```


```{r}
# orient the component so larger = more disadvantaged (positive w/ poverty)
acs$disadvantage <- sign(cor(score[cc], acs$poverty_rate[cc])) * score

predictors <- c("disadvantage","ethnic_hetero","renter_share","youth_share","log_pop")
acs <- acs %>% drop_na(all_of(c("murder_rate", predictors)))
```

```{r}
# collinearity is now controlled?
vif_tab <- sapply(predictors, function(v)
  1/(1 - summary(lm(reformulate(setdiff(predictors, v), v), data = acs))$r.squared))
tibble(variable = predictors,
       vif = round(vif_tab, 2),
       pc1_var_share = ifelse(predictors == "disadvantage",                            round((pc$sdev[1]^2)/sum(pc$sdev^2), 3), NA))
```

## "What correlates with murder across places?" (between metros)

This is a cross-sectional question -- why is metro A more violent than metro B --
so team fixed effects are *dropped* (they would erase exactly the between-metro
variation the question is about). 

**between estimator** that
regresses one row of metro means (the textbook cross-sectional summary), and a
**pooled OLS with year fixed effects and metro-clustered SEs**. 

Predictors are
standardized, so each coefficient is a standardized beta -- directly comparable,
sorted by magnitude to show what is *most* correlated.

```{r correlates-between, message=FALSE, warning=FALSE}
# ---- between estimator: one row per metro (means over years) ----
metro <- acs %>%
  group_by(GEOID) %>%
  summarise(across(all_of(c("murder_rate", predictors)), ~mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  mutate(across(all_of(predictors), ~as.numeric(scale(.x))))   # standardized betas

between_fit <- lm(reformulate(predictors, "murder_rate"), data = metro)

cat("Between estimator --", nrow(metro), "metros (standardized betas):\n")
broom::tidy(between_fit) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate))) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))
```

```{r correlates-pooled, message=FALSE, warning=FALSE}
# ---- pooled OLS + year FE, metro-clustered SEs (keeps within-year variation) ----
pooled <- acs %>% mutate(across(all_of(predictors), ~as.numeric(scale(.x))))
pooled_fit <- lm(reformulate(c(predictors, "factor(year)"), "murder_rate"), data = pooled)
Vp <- sandwich::vcovCL(pooled_fit, cluster = pooled$GEOID, type = "HC1")

cat("Pooled OLS + year FE, metro-clustered SEs (standardized betas):\n")
broom::tidy(lmtest::coeftest(pooled_fit, vcov. = Vp)) %>%
  filter(term %in% predictors) %>%
  arrange(desc(abs(estimate))) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))
```

## "Does changing X move murder within a metro?" (within metros)

Each predictor is standardized by its
**within-metro** SD, so a coefficient is "murders per 100k per one within-metro SD
of the predictor."


```{r correlates-within, message=FALSE, warning=FALSE}
# standardize each predictor by its WITHIN-metro spread (demean by metro, then scale)
z_within <- function(x, g) {
  d <- x - ave(x, g, FUN = function(z) mean(z, na.rm = TRUE))
  d / sd(d, na.rm = TRUE)
}
fe_dat <- acs %>% mutate(across(all_of(predictors), ~z_within(.x, GEOID), .names = "zw_{.col}"))
zw <- paste0("zw_", predictors)

fe_fit <- lm(reformulate(c(zw, "factor(GEOID)", "factor(year)"), "murder_rate"), data = fe_dat)
Vf <- sandwich::vcovCL(fe_fit, cluster = fe_dat$GEOID, type = "HC1")

cat("Metro + year FE, standardized WITHIN betas, metro-clustered SEs:\n")
broom::tidy(lmtest::coeftest(fe_fit, vcov. = Vf)) %>%
  filter(term %in% zw) %>%
  arrange(desc(abs(estimate))) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))
```


disadvantage goes from your strongest clean predictor (β=3.1) to essentially zero (β=0.23, p=.80) once you look within metros over time which cities are violent but a given city getting more or less disadvantaged year-to-year doesn't detectably move its murder rate on this timescale

Between metros, bigger = more murder (+4.4).
Within a metro over time, the coefficient is negative (−0.79, p=.03) 


