---
title: "Year-to-Year Murder Rate vs. Team Performance (within-city z-scores)"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(sandwich)   # cluster-robust covariance
  library(lmtest)     # coeftest() with a supplied vcov
  library(broom)
})

set.seed(123)
options(scipen = 999, digits = 4)
```

## Approach

The question here is narrower than "do teams affect crime": we want to explain a
city's **year-to-year** movement in its murder rate.

- **Outcome:** each metro-year's murder rate, **z-scored within that city**. Every
  city is centered on its own mean and divided by its own SD, so the outcome reads
  as *"how unusual was this year for this city, in that city's own SD units."*
  Because the within-city centering already removes each city's level, **no metro
  fixed effects are used** -- they would be redundant (each city already averages 0).
- **Year fixed effects** are kept, so a season's performance is tested against
  year-to-year movement *beyond* the nationwide crime trend (the 2020 spike, etc.).
- **`win_pct` is combined two ways** -- arithmetic mean and harmonic mean of the
  metro's teams -- and nothing else changes between the two models, so their fit is
  directly comparable. No presence (`has_*`) flags.

Two data issues are handled before any mean is taken:

- **`win_pct` is not on one scale.** MLB's column is a win *count* (19--108, with the
  denominator changing in the 2020 short season) while NFL/NHL/CFB/NBA are 0--1
  fractions. Each league's `win_pct` is therefore **min--max scaled within league**
  onto a common [0, 1] (`wp_s`) before it is combined.
- **A harmonic mean needs strictly positive inputs**, so the scaled values are
  floored at 0.001 (only the single worst team in each league sits at 0).

```{r load-data, message=FALSE, warning=FALSE}
raw <- read_csv("data/metro_acs_model_ready_2010_2023.csv", show_col_types = FALSE) %>%
  mutate(across(c(year, GEOID), as.integer)) %>%
  # The panel is supposed to be one row per (GEOID, year), but 24 metro-years
  # arrive twice -- metros fielding two teams in one league, each row carrying
  # that league's own numbers. Collapse to a single metro-year.
  group_by(GEOID, year) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
            across(!where(is.numeric), first), .groups = "drop")

# --- NBA as a full league ---------------------------------------------------
# The model-ready panel shipped only a `has_NBA` flag, not NBA performance, so
# NBA is built here from its raw files and joined in like the other leagues.
nba_cols <- local({
  cb  <- read_csv("data/metro_acs_2022_cbsa.csv", show_col_types = FALSE) %>%
    dplyr::select(GEOID, NAME) %>% mutate(GEOID = as.integer(GEOID))
  stats <- read_csv("data/NBA/nba_team_stats_2003_2023.csv", show_col_types = FALSE)
  mur   <- read_csv("data/NBA/nba_cities_murder_rates_2003_2026.csv", show_col_types = FALSE)
  apps  <- read_csv("data/NBA/nba_playoff_appearances_since_2003.csv", show_col_types = FALSE)

  panel <- raw %>% distinct(GEOID) %>% left_join(cb, by = "GEOID") %>%
    mutate(city = NAME %>% str_remove(",.*$") %>% str_remove("-.*$"))
  norm  <- function(x) x %>% str_replace("New York City", "New York") %>%
    str_replace("Washington D.C.", "Washington")
  xwalk <- mur %>% distinct(City) %>% mutate(city = norm(City)) %>%
    inner_join(panel, by = "city") %>% dplyr::select(City, city, GEOID)

  team2geo <- mur %>% distinct(`NBA Team`, City) %>% inner_join(xwalk, by = "City") %>%
    dplyr::select(team = `NBA Team`, GEOID)
  city_lut <- xwalk %>% distinct(city, GEOID)
  resolve  <- function(teams) {
    g <- team2geo$GEOID[match(teams, team2geo$team)]
    fb <- map_int(teams, function(t) {
      hit <- str_starts(t, fixed(city_lut$city))
      if (any(hit)) city_lut$GEOID[hit][which.max(nchar(city_lut$city[hit]))] else NA_integer_
    })
    ifelse(is.na(g), fb, g)
  }

  win <- stats %>% mutate(GEOID = resolve(team)) %>% filter(!is.na(GEOID)) %>%
    group_by(GEOID, year) %>% summarise(nba_win_pct = mean(win_loss_perc), .groups = "drop")
  po  <- apps %>% separate_rows(Years, sep = "; ") %>%
    mutate(year = as.integer(Years), GEOID = resolve(Team)) %>% filter(!is.na(GEOID)) %>%
    distinct(GEOID, year) %>% mutate(nba_make_playoffs = 1L)
  mrd <- mur %>% inner_join(xwalk, by = "City") %>% group_by(GEOID, Year) %>%
    summarise(nba_murder_rate = mean(`Murder Rate (per 100k)`), .groups = "drop") %>%
    rename(year = Year)

  win %>%
    full_join(po,  by = c("GEOID", "year")) %>%
    full_join(mrd, by = c("GEOID", "year")) %>%
    mutate(nba_make_playoffs = replace_na(nba_make_playoffs, 0L)) %>%
    arrange(GEOID, year) %>% group_by(GEOID) %>%
    mutate(nba_make_playoffs_prev = lag(nba_make_playoffs)) %>% ungroup()
})

raw <- raw %>% left_join(nba_cols, by = c("GEOID", "year"))
# ---------------------------------------------------------------------------

controls <- c("log_income", "unemployment_rate", "poverty_rate",
              "log_home_value", "log_pop",
              "race_black_nonhispanic_rate", "hispanic_or_latino_rate",
              "bachelors_plus_rate", "pop_18_to_64_rate")

# NOTE: the murder rate is deliberately NOT logged here. It is z-scored within
# city further down, which is the transformation this analysis relies on.
base <- raw %>%
  mutate(
    log_income     = log(median_household_income),
    log_home_value = log(median_home_value),
    log_pop        = log(total_population)
  )
```

```{r stack-panel, message=FALSE, warning=FALSE}
leagues <- c("nfl", "mlb", "nhl", "cfb", "nba")

# one row per (metro, year, team), each league carrying its own murder rate and
# win_pct; then min-max scale win_pct WITHIN league onto a common [0, 1] so the
# leagues can be averaged together (MLB's win count would otherwise swamp them),
# floored at 0.001 so the harmonic mean is defined.
team_long <- map_dfr(leagues, function(lg) {
  base %>%
    transmute(
      GEOID, year, league = toupper(lg),
      murder_rate        = .data[[paste0(lg, "_murder_rate")]],
      win_pct            = .data[[paste0(lg, "_win_pct")]],
      make_playoffs      = .data[[paste0(lg, "_make_playoffs")]],
      make_playoffs_prev = .data[[paste0(lg, "_make_playoffs_prev")]],
      across(all_of(controls))
    ) %>%
    filter(!is.na(murder_rate), !is.na(win_pct))
}) %>%
  group_by(league) %>%
  mutate(wp_s = pmax((win_pct - min(win_pct)) / (max(win_pct) - min(win_pct)), 0.001)) %>%
  ungroup() %>%
  drop_na(all_of(c("wp_s", "make_playoffs", "make_playoffs_prev",
                   "murder_rate", controls)))
```

```{r metro-panel, message=FALSE, warning=FALSE}
# Collapse to one row per (metro, year). The metro's teams are combined two ways
# -- arithmetic and harmonic mean of the scaled win_pct -- on identical rows, so
# the only thing separating the two models below is the combination rule.
metro <- team_long %>%
  group_by(GEOID, year) %>%
  summarise(
    n_teams            = n(),
    murder_rate        = mean(murder_rate),          # raw rate, per 100k
    win_amean          = mean(wp_s),                 # arithmetic mean
    win_hmean          = n() / sum(1 / wp_s),        # harmonic mean
    playoff_share      = mean(make_playoffs),
    playoff_share_prev = mean(make_playoffs_prev),
    across(all_of(controls), first),
    .groups = "drop"
  ) %>%
  # previous-season performance, held IDENTICAL across the two models (the
  # arithmetic lag) so only the contemporaneous encoding differs
  arrange(GEOID, year) %>%
  group_by(GEOID) %>%
  mutate(win_prev = lag(win_amean)) %>%
  ungroup() %>%
  drop_na(win_prev)

# z-score the murder rate WITHIN each city: the outcome becomes "how unusual was
# this year for this city, in that city's own SD units". Cities need at least 3
# years and non-zero variation for a within-city z-score to be defined.
metro <- metro %>%
  group_by(GEOID) %>%
  filter(n() >= 3, sd(murder_rate) > 0) %>%
  mutate(murder_z = as.numeric(scale(murder_rate))) %>%
  ungroup()

tibble(
  observations      = nrow(metro),
  metros            = n_distinct(metro$GEOID),
  years             = n_distinct(metro$year),
  avg_teams_per_obs = round(mean(metro$n_teams), 2),
  outcome_mean      = round(mean(metro$murder_z), 6),   # 0 by construction
  outcome_sd        = round(sd(metro$murder_z), 3)
)
```

## Export the analysis dataset

`metro` is exactly the frame the models below are fit on -- one row per (metro, year),
every variable already transformed. Writing it out means anyone can reproduce these
regressions from the CSV alone, with no access to the raw league files or the NBA
crosswalk. A readable `metro_name` is attached purely as an identifier; it is not a
model variable.

```{r export, message=FALSE, warning=FALSE}
metro_names <- read_csv("data/metro_acs_2022_cbsa.csv", show_col_types = FALSE) %>%
  dplyr::select(GEOID, metro_name = NAME) %>%
  mutate(GEOID = as.integer(GEOID))

final_model_data <- metro %>%
  left_join(metro_names, by = "GEOID") %>%
  dplyr::select(
    # --- identifiers ---
    GEOID, metro_name, year, n_teams,
    # --- outcome ---
    murder_z,            # murder_rate z-scored WITHIN city (the modeled outcome)
    murder_rate,         # raw murders per 100k, mean of the metro's teams' cities
    # --- sports predictors ---
    win_amean,           # arithmetic mean of within-league min-max scaled win_pct
    win_hmean,           # harmonic mean of the same scaled win_pct
    win_prev,            # lag(win_amean) within metro; identical across both models
    playoff_share,       # share of the metro's teams making the playoffs
    playoff_share_prev,
    # --- census controls ---
    all_of(controls)
  ) %>%
  arrange(GEOID, year)

write_csv(final_model_data, "final_model_data.csv")

cat("wrote final_model_data.csv:", nrow(final_model_data), "rows x",
    ncol(final_model_data), "columns\n")
glimpse(final_model_data)
```

## The models

`murder_z` is regressed on the combined win measure, the lagged win measure, the
playoff shares, the census controls, and year fixed effects. **No metro fixed
effects** (the within-city z-score already removed each city's level) and **no
presence flags**.

Model 0 carries **no sports variables at all** -- it is the reference the two
encodings must beat if performance explains year-to-year movement.

```{r fits, message=FALSE, warning=FALSE}
sports_rhs <- paste(c("win_prev", "playoff_share", "playoff_share_prev"), collapse = " + ")
ctrl_rhs   <- paste(c(controls, "factor(year)"), collapse = " + ")

# 0) no sports variables: controls + year FE only
fit_none  <- lm(as.formula(paste("murder_z ~", ctrl_rhs)), data = metro)

# 1) arithmetic mean of the metro's teams' scaled win_pct
fit_amean <- lm(as.formula(paste("murder_z ~ win_amean +", sports_rhs, "+", ctrl_rhs)),
                data = metro)

# 2) harmonic mean of the metro's teams' scaled win_pct
fit_hmean <- lm(as.formula(paste("murder_z ~ win_hmean +", sports_rhs, "+", ctrl_rhs)),
                data = metro)

# SEs clustered by metro: a metro's years are correlated
clustered <- function(fit, term) {
  V <- sandwich::vcovCL(fit, cluster = metro$GEOID, type = "HC1")
  broom::tidy(lmtest::coeftest(fit, vcov. = V)) %>% filter(term == !!term)
}
```

Sanity check that `final_model_data.csv` really is sufficient: refit model 1 from the
CSV alone and confirm it reproduces `fit_amean` exactly.

```{r export-reproduce, message=FALSE, warning=FALSE}
check <- read_csv("final_model_data.csv", show_col_types = FALSE)

refit <- lm(as.formula(paste("murder_z ~ win_amean +", sports_rhs, "+", ctrl_rhs)),
            data = check)

tibble(
  n_match            = nrow(check) == nrow(metro),
  coefficients_match = isTRUE(all.equal(coef(refit), coef(fit_amean))),
  r_squared_match    = isTRUE(all.equal(summary(refit)$r.squared,
                                        summary(fit_amean)$r.squared))
)
```

## First: the year dummies already explain almost everything

A raw R^2 from these models is badly misleading, and it is worth seeing why before
reading any of them. Once each city is standardized against **its own** mean and SD,
every city's murder trajectory collapses onto essentially the *same national curve*
(flat through the 2010s, the 2020 spike, the 2021--22 aftermath). So `factor(year)`
on its own accounts for the overwhelming majority of the within-city variance, and
only a small idiosyncratic remainder is left for anything else to explain.

```{r year-dominates, message=FALSE, warning=FALSE}
r2 <- function(f) summary(f)$r.squared
tibble(
  specification = c("murder_z ~ factor(year) only",
                    "murder_z ~ controls only",
                    "murder_z ~ controls + factor(year)  (model 0)"),
  r_squared = c(r2(lm(murder_z ~ factor(year), data = metro)),
                r2(lm(as.formula(paste("murder_z ~", paste(controls, collapse = " + "))),
                      data = metro)),
                r2(fit_none))
) %>% mutate(r_squared = round(r_squared, 4))
```

## How much year-to-year variance is explained?

Because the outcome is a within-city z-score, `r_squared` is the share of a city's
year-to-year variance explained. Given the table above, the *level* of R^2 is
uninformative -- what matters is the **gain over model 0** (`r2_gain`), i.e. what the
sports terms add once the national year pattern and the census controls are already in.

```{r variance-table, message=FALSE, warning=FALSE}
mse <- function(f) mean(residuals(f)^2)

tibble(
  model = c("0. no sports variables (controls + year FE)",
            "1. arithmetic mean of win_pct",
            "2. harmonic mean of win_pct"),
  n            = nrow(metro),
  n_predictors = c(sum(!is.na(coef(fit_none))),
                   sum(!is.na(coef(fit_amean))),
                   sum(!is.na(coef(fit_hmean)))),
  r_squared    = c(summary(fit_none)$r.squared,
                   summary(fit_amean)$r.squared,
                   summary(fit_hmean)$r.squared),
  # what the sports block actually buys over the no-sports baseline
  r2_gain      = c(0,
                   summary(fit_amean)$r.squared - summary(fit_none)$r.squared,
                   summary(fit_hmean)$r.squared - summary(fit_none)$r.squared),
  adj_r_squared = c(summary(fit_none)$adj.r.squared,
                    summary(fit_amean)$adj.r.squared,
                    summary(fit_hmean)$adj.r.squared),
  MSE          = c(mse(fit_none), mse(fit_amean), mse(fit_hmean))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 5)))
```

A formal test of the whole sports block (the win measure, its lag, and both playoff
shares) against the no-sports baseline:

```{r sports-block-ftest, message=FALSE, warning=FALSE}
bind_rows(
  broom::tidy(anova(fit_none, fit_amean))[2, ] %>% mutate(model = "arithmetic mean"),
  broom::tidy(anova(fit_none, fit_hmean))[2, ] %>% mutate(model = "harmonic mean")
) %>%
  dplyr::select(model, df, sumsq, statistic, p.value) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## The win coefficients (metro-clustered SEs)

Each coefficient is the effect of the combined win measure on the city's
**within-city z-scored** murder rate: a coefficient of `b` means a 1-unit move in
the [0, 1] combined win measure shifts that city's murder rate by `b` of its own
standard deviations.

```{r win-coefs, message=FALSE, warning=FALSE}
bind_rows(
  clustered(fit_amean, "win_amean") %>% mutate(model = "arithmetic mean"),
  clustered(fit_hmean, "win_hmean") %>% mutate(model = "harmonic mean")
) %>%
  dplyr::select(model, term, estimate, std.error, statistic, p.value) %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))
```

## Model summaries

Year dummies are suppressed so the win, playoff, and census coefficients are
readable. Both models still contain `factor(year)`.

```{r summary-helper, message=FALSE, warning=FALSE}
summary_hide_fe <- function(fit) {
  s <- summary(fit)
  s$coefficients <- s$coefficients[!grepl("^factor\\(", rownames(s$coefficients)), ,
                                   drop = FALSE]
  if (!is.null(s$aliased)) s$aliased <- s$aliased[!grepl("^factor\\(", names(s$aliased))]
  s
}
```

```{r summary-none}
# 0. no sports variables
summary_hide_fe(fit_none)
```

```{r summary-amean}
# 1. arithmetic mean of win_pct
summary_hide_fe(fit_amean)
```

```{r summary-hmean}
# 2. harmonic mean of win_pct
summary_hide_fe(fit_hmean)
```

## Reading the results

- The outcome is each city's murder rate expressed in **its own** standard
  deviations, so a coefficient is a within-city effect by construction -- stable
  differences between cities (big cities have teams *and* more crime) cannot drive
  it, which is why no metro fixed effects are needed.
- **Do not read the raw R^2.** `factor(year)` alone explains ~91% of the within-city
  variance, because standardizing each city against its own mean and SD makes every
  city trace the same national curve. The number that carries information is
  `r2_gain`: what the sports block adds *on top of* the year pattern and the controls.
  The block F-test above is the formal version of the same question.
- The arithmetic and harmonic means are computed on **identical rows** from the same
  scaled `wp_s`, so any difference between models 1 and 2 is attributable purely to
  the combination rule (the harmonic mean punishes a metro with one very bad team
  much harder than the arithmetic mean does).
